---
layout: post
title: dplyr is pretty good.
comments: true
date: 2014-2-11 15:20:00 
author: [{name: "Joe", url: "jofrhwld.github.io"}]
---

R man-god Hadley Wickham is at it again, releasing the ```dplyr``` package.
It's similar to his previous ```plyr``` package, but designed specifically for working with data frames, and highly optimized.
Just to show off a nice feature of its functionality, here's how we'd do the calcuation of TD-Deletion rates as I orignally described [here](http://val-systems.blogspot.co.uk/2012/05/on-calculating-exponents.html).

First, let's load the data.

```{r cache = T}
library(plyr)
library(RCurl)
b <- read.delim(textConnection(getURL("https://raw.github.com/JoFrhwld/TD-Classifier/master/Data/new_buckeye.txt")))
eval(parse(text = getURL("https://raw.github.com/JoFrhwld/TD-Classifier/master/R/buckCoder.R")))
 
b <- subset(b, !(FolSeg %in% c("apical")) & PreSeg != "/r/")
 
psm <- subset(b, Gram2 %in% c("past","semiweak","mono"))
 
psm$Word <- as.character(psm$Word)
 
psm$Gram2 <- as.factor(psm$Gram2)
psm$Gram2 <- relevel(psm$Gram2, "past")
```


The first key ```dplyr``` function is ```group_by()```, which creates a grouped data frame (it also worth a bunch of other data table types, including ```data.table```, see ```?group_by()```).
```{r}
library(dplyr)
psm_gr <- group_by(psm, Gram2, Speaker, Word)
psm_gr
```

Now, first things first, there are ```r nrow(psm_gr)``` rows to this data, and I made the previously stupid move of just printing it all, but the ```grouped_df``` class is a bit smarter in the way it prints, just showing the first few rows and columns. That's pretty nice.
I'll also call your attention to the second line of the printed output, where it says ```Groups: Gram2, Speaker, Word```.

Next, we can ```summarise``` the data.

```{r}
summarise(psm_gr, td = mean(td))
```

Here we see the mean rate of TD retention per grammatical class, per speaker, per word. 
But look at the second line of the output.
It started out with ```Groups: Gram2, Speaker, Word```, but now it's ```Groups: Gram2, Speaker```.
Apparently, ever time you ```summarise``` a grouped data frame, the outermost layer of grouping is stripped away.
This is really useful for the kind of nested calculation of proportions that I argued is necessary for descriptive statistics in that Val Systems post.
And it's easy to do with ```dplyr``` which uses a custom operator, ```%.%```, which allows you to chain up these data manipulation operations.
Here's how you'd do the calculation of the rate of TD retention.


```{r tidy = F}
td_gram <- psm %.% 
            group_by(Gram2, Speaker, Word) %.% 
            summarise(td = mean(td)) %.% # over word
            summarise(td = mean(td)) %.% # over speaker
            summarise(td = mean(td))     # over grammatical class
td_gram
```

I've been doing something similar for calculating mean F1 and F2 for vowels, which would look like this.

```{r tidy = F, eval=F}
vowel_means <- vowels  %.%
                group_by(Speaker, Age, Sex, DOB, VowelClass, Word) %.%
                summarise(F1.n = mean(F1.n), F2.n = mean(F2.n)) %.%  
                summarise(F1.n = mean(F1.n), F2.n = mean(F2.n))
```