---
title: "Github Classroms - Revisted in 2026"
format: html
editor: visual
draft: true
---

I've made a few posts before about using GitHub Classrooms for teaching quant/comp courses.

-   [Some lessons in using Github Classroom and Codespaces](/posts/2024/02/2024-02-14_gh-classroom/) (February 2024)

-   [Troubleshooting RStudio for GitHub Classrom and Codespaces](/posts/2024/09/2024-09-06_rserver-codespaces/) (September 2024)

I've gotten a few more GH Classroom courses under my belt, so I thought I'd share a bit more about my workflow and configuration.

```{r}
library(distributional)
library(tidyverse)
library(ggdist)
```

```{r}
r = 0.9
dist <- dist_multivariate_normal(
  mu = list(c(0,0)), 
  sigma = list(matrix(c(1,r,r,1), ncol=2)))
dimnames(dist) <- c("x", "y")

generate(dist, 10000) |> 
  pluck(1) |> 
  as_tibble() ->
  samp

samp |> 
  ggplot(
    aes(x, y)
  ) + 
  geom_point()+
  coord_fixed()

samp |> 
  summarise(
    across(x:y, sd, .names = "{.col}_sd"),
    r = cor(x,y)
  )

```

```{r}
x <- c("kitten", "sitting", "knittig")
adist(x, x)
```
