---
title: "Introducing tidynorm"
description: |
  Here's a brief introduction to the new `{tidynorm}` package.
date: 2025-06-16
format: html
editor: visual
knitr: 
  opts_chunk: 
    crop: true
    fig.width: 6
    fig.asp: 0.618
    out.width: 70%
    fig.align: center
bibliography: references.bib
---

```{r}
#| echo: false
library(tidynorm)
```

::::::: callout-tip
## The upshot

The `{tidynorm}` package has convenience functions for normalizing

:::::: {layout-ncol="3"}
<div>

Point measurements

```{r}
#| echo: false
#| results: asis
ns <- apropos("^norm_[^tg][^(dc)].*")
for(x in ns){
  print(glue::glue("- `{x}()`"))
}
```

</div>

<div>

Formant Tracks

```{r}
#| echo: false
#| results: asis
ns <- apropos("^norm_track_[^g].*")
for(x in ns){
  print(glue::glue("- `{x}()`"))
}
```

</div>

<div>

DCT coefficients

```{r}
#| echo: false
#| results: asis
ns <- apropos("^norm_dct_[^g].*")
for(x in ns){
  print(glue::glue("- `{x}()`"))
}
```

</div>
::::::

As well as generic functions to implement your own normalization method.

You can install `{tidynorm}` in your preferred way from CRAN.

```{r}
#| eval: false
install.packages("tidynorm")
```
:::::::

```{r}
#| message: false
#| code-fold: true
#| code-summary: Loading Packages
library(tidynorm)
library(tidyverse)
library(scales)
library(ggtext)
library(gt)
source(here::here("_defaults.R"))
```

## What is speaker vowel normalization?

Imagine a very tall person from London speaking to you.
You can probably imagine what their accent sounds like.
Now imagine a very short person speaking to you in the same accent.
In reality, if you heard these two people speaking in what you perceive to be identical accents, the acoustics of their speech will be different due to their (likely) different vocal tract lengths (VTL).

Using some rough heuristics and assumptions, the overall vowel spaces of these two speakers might look something like this:

```{r}
#| label: vowel-polygon
#| code-fold: true
#| code-summary: "guestimates functions"

vtl_2_formant <- function(vtl, f = 1){
  dF = 34300/(2*vtl)
  dF * (f * 0.5)
}

vowel_polygon <- function(F1, F2){
  tibble(
    F1 = c(
      F1 * 0.6,
      F1 * 1.45,
      F1 * 0.6
      ),
    F2 = c(
      F2 * 1.5,
      F2,
      F2 * 0.52
    )
  )
}
```

```{r}
#| code-fold: true
#| code-summary: vowel-space-plot
#| renderings: [light, dark]
tibble(
  vtl = seq(14, 17, length = 2)
) |> 
  mutate(
    F1 = vtl_2_formant(vtl, 1),
    F2 = vtl_2_formant(vtl, 2),
    F3 = vtl_2_formant(vtl, 3)
  ) |>  mutate(
    vtl = factor(vtl, labels = c("short", "long"))
  ) ->
  speaker_formants

speaker_formants |> 
  reframe(
    .by = vtl,
    vowel_polygon(F1, F2)
  ) |> 
  ggplot(
    aes(F2, F1)
  ) +
  geom_polygon(
    aes(group = vtl, fill = vtl, color = vtl),
    linewidth = 1,
    alpha = 0.6
  ) +
  scale_y_reverse() +
  scale_x_reverse() +
  labs(fill = "VTL", color = "VTL") +
  coord_fixed() -> p

p
p+theme_dark()
```

These speakers' overall vowel spaces have different center points, and cover different areas.

```{r}
#| code-fold: true
#| code-summary: plotting code

speaker_formants |> 
  ggplot(
    aes(F2, F1)
  ) +
  geom_point(
    aes(color = vtl),
    size = 5
  ) +
  scale_y_reverse() +
  scale_x_reverse() +
  labs(title = "vowel space center") -> center_p

speaker_formants |> 
  reframe(
    .by = vtl,
    vowel_polygon(F1, F2)
  ) |> 
  summarise(
    .by = vtl,
    b = diff(range(F1)),
    h = diff(range(F2))
  ) |> 
  mutate(
    a = (b/100 * h/100)*2
  ) |> 
  ggplot(
    aes(
      vtl, a
    )
  )+
  geom_col(aes(fill = vtl)) +
  labs(
    y = "area (kHz<sup>2</sup>)",
    title = "vowel space area"
  )+
  theme(
    axis.title.y = element_markdown()
  ) ->
  area_plot
```

::: light-content
```{r}
#| echo: false
#| layout-ncol: 2
#| crop: false

center_p
area_plot
```
:::

::: dark-content
```{r}
#| echo: false
#| crop: false 
#| layout-ncol: 2
center_p + theme_dark()
area_plot + theme_dark()

```
:::

The goal of any speaker vowel normalization method is to try to line up speakers' vowel spaces so that speaker A's highest, frontest vowels are lined up with speaker B's, so that speaker A's lowest, backest vowels are lined up with speaker B's, etc.
Once we have their vowel spaces aligned in a such way that we know similarities between them are matched, we can start investigating any differences.

::: callout-tip
## Vowel Space ≠ Pitch

One really important thing to keep in mind is that vowel space differences due to different vocal tract lengths are *not* the same as differences in speakers' pitch.
Differences in speakers' pitch are caused by differences in how their vocal folds vibrate.
You can have two speakers' with the same exact pitch, but very different vocal tract lengths (& vowel spaces), and vice versa.
:::

## Normalization methods

All normalization methods involve some kind of shift in the location of a speaker's vowel space by some value $L$, scaling the size of a speaker's vowel space by some value $S$, or both.

$$
F' = \frac{F-L}{S}
$$

They way normalization methods mainly differ is in terms of

-   Transformations applied to the original formant values before calculating $L$ and $S$ (e.g. log, bark).

-   The exact functions used to calculate $L$ and $S$ (e.g. mean, standard deviation)

-   The scope over which $L$ and $S$ are calculated (e.g. across all formants at once, or one formant at a time).

For example, the logic behind Nearey normalization [@neareyPhoneticFeatureSystems1978] is that after log transforming vowel spaces, they should really only differ in terms of the location of the centers, not in terms of their area.

```{r}
#| code-fold: true
#| code-summary: plotting code
#| renderings: [light, dark]

speaker_formants |> 
  reframe(
    .by = vtl,
    vowel_polygon(F1, F2)
  ) |> 
  mutate(across(F1:F2, log)) |> 
  ggplot(
    aes(F2, F1)
  ) +
  geom_polygon(
    aes(
      group = vtl, 
      fill = vtl, 
      color = vtl
      ),
    alpha = 0.6,
    linewidth = 1
  ) +
  scale_y_reverse("log(F1)") +
  scale_x_reverse("log(F2)") +
  coord_fixed() -> p

p
p+theme_dark()
```

So what Nearey normalization does is

-   log transform the data

-   take the average value across *all* formants

-   subtracts that value from each token

So where $i$ is the formant number, and $j$ is the token number:

$$
F_{ij}' = \frac{\log F_{ij}-L}{1}
$$

$$
L = \frac{1}{MN} \sum_{i=1}^3\sum_{j=1}^N \log F_{ij}
$$

It's possible to do this yourself using [some tidyverse verbs](https://lingmethodshub.github.io/content/R/tidy-norm/), but it involves some pivoting between wide and long.
This, combined with [my work on normalizing vowel formant tracks](https://jofrhwld.github.io/dct_normalization/) motivated me to create the `{tidynorm}` package.

## Normalizing with `{tidynorm}`

```{r}
library(tidynorm)
library(ggdensity)
```

Let's start with two speakers' unnormalized data

```{r}
#| code-fold: true
#| code-summary: plotting code
#| renderings: 
#|   - light
#|   - dark
speaker_data |> 
  ggplot(
    aes(F2, F1, color = speaker)
  ) +
  stat_hdr(
    probs = 0.95,
    aes(fill = speaker),
    linewidth = 1
  )+
  scale_x_reverse() +
  scale_y_reverse() +
  guides(
    alpha = "none"
  )+
  coord_fixed()->p_unnorm

p_unnorm
p_unnorm + theme_dark()
```

```{r}
#| echo: false
speaker_data |> 
  gt_preview() |> 
  opt_table_font(
    "Public Sans"
  ) |> 
  fmt_number(
    decimals = 0, use_seps = F
  )
```

We can implement the logic of Nearey normalization in tidynorm's function `norm_generic()`.

```{r}
#| message: true
speaker_nearey <- speaker_data |>
  norm_generic(
    # the formants to normalize
    F1:F3,
    
    # provide the speaker id column
    .by = speaker,
    
    # pre calculation transformation function
    .pre_trans = log,
    
    # location calculation
    .L = mean(.formant, na.rm = T),
    
    # scope
    .by_formant = F,
    .by_token = F
  )
```

I tried to include a helpful message describing what kind of normalization just happened.
Here's how the normalized data looks.

```{r}
#| code-fold: true
#| code-summary: plotting code
#| renderings: 
#|   - light
#|   - dark
speaker_nearey |> 
  ggplot(
    aes(F2_n, F1_n, color = speaker)
  ) +
  stat_hdr(
    probs = 0.95,
    aes(fill = speaker),
    linewidth = 1
  )+
  scale_x_reverse() +
  scale_y_reverse() +
  guides(
    alpha = "none"
  )+
  coord_fixed()->p_nearey

p_nearey
p_nearey + theme_dark()
```

### Implementing Lobanov

The Lobanov normalization technique [@lobanov] essentially z-scores each formant ($L$ = the mean, $S$ = the standard deviation).
We can see how that logic can be implemented in `norm_generic()` as well.

```{r}
#| message: true
speaker_lobanov <- speaker_data |> 
  norm_generic(
    # the formants to normalize
    F1:F3,
    
    # provide the speaker id column
    .by = speaker,
    
    # location calculation
    .L = mean(.formant, na.rm = T),
    
    # scale calculation
    .S = sd(.formant, na.rm = T),
    
    # scope
    .by_formant = T,
    .by_token = F
  )
```

```{r}
#| code-fold: true
#| code-summary: plotting code
#| renderings: 
#|   - light
#|   - dark
speaker_lobanov |> 
  ggplot(
    aes(F2_n, F1_n, color = speaker)
  ) +
  stat_hdr(
    probs = 0.95,
    aes(fill = speaker),
    linewidth = 1
  )+
  scale_x_reverse() +
  scale_y_reverse() +
  guides(
    alpha = "none"
  )+
  coord_fixed()->p_lobanov

p_lobanov
p_lobanov + theme_dark()

```

### Convenience functions

Instead of needing to write out the centering and scaling functions yourself every time, I've included convenience functions for some established normalization methods, including

-   `norm_lobanov()`

-   `norm_nearey()`

-   `norm_deltaF()`

-   `norm_wattfab()`

-   `norm_barkz()`

They're all just wrappers around `norm_generic()`, so if you're looking for some inspiration implementing your own normalization method, have a look at [the source](https://github.com/JoFrhwld/tidynorm/blob/main/R/norm.R) to see how I implemented these.

We can apply multiple normalization methods to the same data set by chaining them.

```{r}
speaker_multi <- speaker_data |> 
  norm_nearey(
    F1:F3,
    .by = speaker,
    .silent = TRUE
  ) |> 
  norm_lobanov(
    F1:F3,
    .by = speaker, 
    .silent = TRUE
  ) |> 
  norm_deltaF(
    F1:F3,
    .by = speaker, 
    .silent = T
  )

```

If you've lost track of which normalization methods you've used, and where the normalized values have gone, you can print an information message with `check_norm()`.

```{r}
#| message: true
check_norm(speaker_multi)
```

## Normalizing Formant Tracks

While I think [the advice I have for normalizing formant tracks](https://jofrhwld.github.io/dct_normalization/) is good, I admit it's fairly complex.
So I've also implemented formant track normalization methods:

-   `norm_track_generic()`

-   `norm_track_lobanov()`

-   `norm_track_nearey()`

-   `norm_track_wattfab()`

-   `norm_track_barkz()`

Let's look at one of them in action on formant track data.

```{r}
#| renderings: [light, dark]
#| code-fold: true
#| code-summary: plotting code
speaker_tracks |>   
  filter(
    .by = c(speaker, id),
    !any(F1 > 1200)
  ) ->
  speaker_tracks

speaker_tracks |> 
  ggplot(
    aes(F2, F1, color = speaker)
  )+
  geom_path(
    alpha = 0.2,
    aes(group = interaction(speaker, id))
  )+
  guides(
    color = guide_legend(override.aes = list(alpha = 1))
  )+
  scale_x_reverse()+
  scale_y_reverse() -> p
p
p + theme_dark()
```

```{r}
#| echo: false

speaker_tracks |> 
  gt_preview() |> 
  opt_table_font(
    "Public Sans"
  ) |> 
  fmt_number(
    columns = starts_with("F"),
    decimals = 0, use_seps = F
  ) |> 
  fmt_number(
    columns = t,
    decimals = 2
  )
```

In addition to identifying the speaker ID column, we also need to provide a column that uniquely identifies each token (in this data set, `id`) and we *can* provide an optional column of time information.[^1]

[^1]: By default, these track normalization methods will also slightly smooth the formant tracks.
    But if you don't want that, you can pass it `.order = NA.`

```{r}
#| message: true
speaker_track_lobanov <- speaker_tracks |> 
  norm_track_lobanov(
    # the formants to normalize
    F1:F3,
    
    # provide the speaker id column
    .by = speaker,
    
    # provide the token id column
    .token_id_col = id,
    
    # provide a time column
    .time_col = t
  )
```

```{r}
#| renderings: [light, dark]
#| code-fold: true
#| code-summary: plotting code
speaker_track_lobanov |> 
  ggplot(
    aes(F2_z, F1_z, color = speaker)
  )+
  geom_path(
    alpha = 0.2,
    aes(group = interaction(speaker, id))
  )+
  guides(
    color = guide_legend(override.aes = list(alpha = 1))
  )+
  scale_x_reverse()+
  scale_y_reverse() -> p

p
p + theme_dark()
```

## Extending `{tidynorm}`

If there is a normalization method you really like, or are just interested in, and aren't sure how to implement it in `{tidynorm}`, add an issue (ideally with a reference or some math) on the [github issues page](https://github.com/JoFrhwld/tidynorm/issues).

## Closing thoughts

This was a complex, but really enjoyable package to write.
In addition to wrapping my head around ["tidy evaluation"](https://adv-r.hadley.nz/evaluation.html), there was a lot of conceptual work in figuring out how to implement one consistent data processing workflow in `norm_generic()` that could carry out the normalization methods that have been described in the literature, each in their own way.
Something like

> Take each formant column and z-score it.

is pretty straightforward, but something like

> Transform Hz into Bark, then for each token, subtract the third formant from the first and second formants.

is a little trickier to include in the same workflow.

The `{tidynorm}` method most different from the method as described in the literature is `norm_wattfab()`.
As described by @wattEvaluationTechniqueImproving2002, the method involves calculating the means of corner vowels.
Doing that inside of a tidynorm workflow isn't impossible, but would sacrifice a lot of generality, and would require users to provide a vowel-class column name every time (since I can't assume what everyone's data columns are called).
Instead, I went for the shortcut method also used by @johnsonDFMethodVocal2020, and calculated their $S$ values based on the mean across each formant.

I might revisit that in the future, but it would require a much more hands-on approach from the user than the other convenience functions currently do.
