---
title: "Light vs Dark &lt;color&gt;"
format: html
date: 2025-07-14
editor: visual 
code-fold: true
knitr: 
  opts_chunk: 
    message: false
    warning: false
bibliography: references.bib
---

So now I can [finally](../2025-07-09_color-survey) get to [visualizing](../2025-07-12_color-sphere) the effect of "light" and other modifiers on colors!
When I eventually get to the `{plotly}` code, there's nothing tidy going on, so I'll be code-folding most of this stuff.

```{r}
#| code-fold: false
source(here::here("_defaults.R"))
library(tidyverse)
library(tidytuesdayR)
library(tinytable)
library(plotly)
library(ggdist)

library(patchwork)
set.seed(2025-07-08)
```

```{r}
max_text_contrast <- function(bgs){
  w_contrast <- abs(
    contrast_ratio(
      "white", bgs,
      algorithm = "APCA"
    )[,1]
  )
  b_contrast <- abs(
    contrast_ratio(
      "black", bgs,
      algorithm = "APCA"
    )[,1]
  )
  out_cols <- c("white", "black")

  out_cols[
    cbind(w_contrast, b_contrast) |>
      apply(1, which.max)
  ]
  
}
```

```{r}
#| code-fold: false
#| eval: false
# eval: false
# downloading & saving to avoid 
# downloading on every quarto render
tuesdata <- tidytuesdayR::tt_load('2025-07-08')

fs::dir_create("data")
write_rds(tuesdata, "data/tuesdata.rds")
```

```{r}
#| code-fold: false
tuesdata <- read_rds("data/tuesdata.rds")

answers <- tuesdata$answers
color_ranks <- tuesdata$color_ranks
users <- tuesdata$users
```

```{r}
#| code-fold: false
color_ranks |> 
  pull(hex) |>
  hex2RGB() |> 
  as("HLS") |> 
  slot("coords") |> 
  as_tibble() |> 
  bind_cols(
    color_ranks
  ) ->
  color_hls
```

## The goal

Let's take "blue" and "light blue" as a quick example.
Here are these two labels the "average" hex code associated with these labels, as well as the Hue, Lightness, Saturation values.

```{r}
color_hls |> 
  filter(
    color %in% c("blue", "light blue")
  ) |> 
  mutate(
    hls = str_glue(
      "HLS({round(H)}, {round(L, digits = 2)}, {round(S, digits = 2)})"
    )
  ) |> 
  select(color, hex, hls) ->
  blues

blues |> 
  tt() |> 
  style_tt(
    i = 1:2,
    j = 2,
    background = blues$hex,
    color = max_text_contrast(blues$hex)
  )
```

Unsurprisingly, blue ➝ light blue involves an increase in the lightness (from 0.44 to 0.79), but there's also about a 20 degree rotation of the hue towards green.
The saturation stays about constant, but that's not true for every "\<color\>", "light \<color\>" pair.
"Indigo" desaturates quite a bit when it becomes "light indigo."

```{r}
color_hls |> 
  filter(
    color %in% c("indigo", "light indigo")
  ) |> 
  mutate(
    hls = str_glue(
      "HLS({round(H)}, {round(L, digits = 2)}, {round(S, digits = 2)})"
    )
  ) |> 
  select(color, hex, hls) ->
  indigos

indigos |> 
  tt() |> 
  style_tt(
    i = 1:2,
    j = 2,
    background = indigos$hex,
    color = max_text_contrast(indigos$hex)
  )
```

The path these colors follow when they move from "\<color\>" to"\<modifier\> \<color\>" is what I want to visualize.

## Modifiers

### Single word terms

First of all, I need to identify every one word color term so that I can go on to identify their modified versions.
This involves some regex work

```{r}
#| code-fold: false
color_hls |> 
  filter(
    # filter out any labels
    # containing a space, slash or hyphen
    str_detect(color, "[ /-]", negate = T),
    # forcing this by hand
    !color %in% c("dark", "pale")
  ) ->
  single_terms

nrow(single_terms)
```

Let's get a look at 5 randomly sampled terms.

```{r}
single_terms |> 
  slice_sample(n = 5) |> 
  select(color, rank, hex) |> 
  arrange(rank) ->
  example_single

example_single |> 
  tt() |> 
  style_tt(
    i = 1:5,
    j = 3,
    background = example_single$hex,
    color = max_text_contrast(example_single$hex)
  )
```

I don't study color terms, and am only vaguely aware of the debates in the area.
My shallow investigation into it turns up a 1969 book, *Basic color terms: their universality and evolution* by Berlin & Kay (third edition: @berlinBasicColorTerms1999), that seems to be a touchstone work.
I couldn't quickly get an electronic copy, but I was able to find a 2001 encyclopedia entry by Kay [@kayColorTermsLinguistics2001] that includes in the definition of "basic color term"

-   must be monolexemic

-   must not be the name of a kind of object

So from the 5 sampled color terms here, I think these criteria would exclude "blurple" for being a portmanteau, and "poop", "grapefruit" and "bruise" for being some kind of object.
I'm still going to treat all of them as single color terms for this post though.

I'll concatenate all single color words into a big regular expression.

```{r}
#| code-fold: false
single_terms |> 
  pull(color) |> 
  str_c(collapse = "|") ->
  single_word_regex

color_hls |> 
  mutate(
    n_color = str_count(
      color, single_word_regex
    )
  ) ->
  color_hls
```

Let's get a sense of how many color terms show up in the labels.

```{r}
color_hls |> 
  count(
    n_color
  ) |>
  tt() |>
  style_tt(
    j = 2, align = "r"
  )
```

The nine color labels that had 0 matches for any of the single color terms was interesting.

```{r}
color_hls |> 
  filter(n_color == 0) ->
  zeros

zeros |> 
  select(
    color, rank, hex
  ) |> 
  tt(caption = "0 single color words") |> 
  style_tt(
    i = 1:nrow(zeros),
    j = 3,
    background = zeros$hex,
    color = max_text_contrast(zeros$hex)
  )
```

-   Two of them ("pale" and "dark") are modifiers used on their own that I specifically excluded from the single color terms because I wanted to investigate them as modifiers.

-   Four of them ("terra cotta", "robin's egg", "egg shell", and "macaroni and cheese") are multi-word descriptions that contain no words that were used on their own as color terms (although, the alternative spelling "terracotta" was included).

-   The remaining three ("light urple", "burnt siena", and "light lavendar") misspell the single color terms that they contain.

It also seems interesting that the maximum number of color terms included in a label is 2!
There weren't any "blueish green yellow" or anything.

### Getting the Modifiers

I'll be identifying modified colors labels as

-   labels that include just 1 color term,

-   and have just 2 words,

-   and the second word is a color term

And I'll be defining "modifiers" as the first word in these labels.

```{r}
#| code-fold: false
color_hls |> 
  filter(
    # 1 color term
    n_color == 1,
    # 2 words
    str_count(color, " ") == 1,
    # second word is color
    str_detect(
      color,
      str_c(" ", single_word_regex,"$")
    )
  ) |> 
  separate_wider_delim(
    color,
    delim = " ",
    names = c("modifier", "base")
  ) ->
  modified_hls
```

Let's get a look at the most common modifiers

```{r}
modified_hls |> 
  count(modifier) |> 
  arrange(desc(n)) |> 
  slice_head(n = 6) ->
  top_mod

top_mod |> 
  tt() |> 
  style_tt(
    j = "n",
    align = "r"
  )
```

I'll also need to get the base color data for making the comparisons.

```{r}
#| code-fold: false
color_hls |> 
  filter(
    n_color == 1,
    str_count(color, " ") == 0
  )  |> 
  rename(
    base = color
  ) ->
  base_colors
```

And then a join will get me the modifier, base color in one data frame.

```{r}
#| code-fold: false
modified_hls |> 
  left_join(
    base_colors,
    by = "base",
    suffix = c("_mod", "_base")
  ) |> 
  filter(is.finite(L_base)) ->
  comp_hls
```

## General modifier analysis

The plotly code gets gnarly, so this is all going to be folded.

```{r}
make_mod_segments <- function(comp_hls, mod) {
  comp_hls |> 
    filter(
      modifier == mod
    ) |> 
    mutate(
      H_mod = case_when(
        H_mod - H_base > 180 ~ H_mod - 360,
        H_mod - H_base < -180 ~ 360 + H_mod,
        .default = H_mod
      )
    ) |> 
    reframe(
      .by = base,
      H = seq(H_base, H_mod, length = 50),
      L = seq(L_base, L_mod, length = 50),
      S = seq(S_base, S_mod, length = 50),
      point = 1:50
    ) |> 
    mutate(
      hex = HLS(H, L, S) |> hex(),
      H_radian = H * (pi/180),
      z = (L*2) - 1,
      max_S = sqrt(abs((z^2) - 1)),
      weighted_S = S * max_S,
      x = sin(H_radian) * weighted_S,
      y = cos(H_radian) * weighted_S,
      color = base
    ) |> 
    nest(
      .by = base,
      data = c(x, y, z, hex, color)
    ) 
}

base_sphere <- function(){
  plot_ly(
    type = "scatter3d",
    mode = "lines"
  ) |>   
  add_trace(
    x = sin(seq(0, pi*2, length = 100)),
    y = cos(seq(0, pi*2, length = 100)),
    z = rep(0, 100),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  add_trace(
    x = sin(seq(0, pi*2, length = 100)),
    z = cos(seq(0, pi*2, length = 100)),
    y = rep(0, 100),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  add_trace(
    y = sin(seq(0, pi*2, length = 100)),
    z = cos(seq(0, pi*2, length = 100)),
    x = rep(0, 100),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  add_trace(
    x = sin(seq(0, pi*2, length = 100)) * 0.5,
    y = cos(seq(0, pi*2, length = 100)) * 0.5,
    z = rep(0, 100),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  add_trace(
    x = sin(seq(0, pi*2, length = 100)) * 0.5,
    z = cos(seq(0, pi*2, length = 100)) * 0.5,
    y = rep(0, 100),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  add_trace(
    y = sin(seq(0, pi*2, length = 100)) * 0.5,
    z = cos(seq(0, pi*2, length = 100)) * 0.5,
    x = rep(0, 100),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  add_trace(
    x = rep(0, 5),
    y = rep(0, 5),
    z = seq(-1, 1, length = 5),
    mode = "lines",
    line = list(color = "grey"),
    hoverinfo = "none"
  ) |> 
  layout(
    showlegend = F,
    autosize = F,
    scene = list(
      xaxis = list(
        domain = c(-1, 1),
        color = "white"
      ),
      yaxis = list(
        domain = c(-1, 1),
        color = "white"
      ),
      zaxis = list(
        domain = c(-1, 1),
        color = "white"
      )      
    )
  ) 
}

make_mod_plot <- function(comp_hls, mod, mode  = "light") {
  
  segments <- make_mod_segments(comp_hls, mod)
  
  p <- base_sphere()
  
  for(i in seq_along(segments$data)){
    d <- segments$data[[i]]
    col <- segments$base[i]
    markers <- c(rep("circle", length(d$x)-1), "diamond")
    sizes <- c(10, rep(1, length(d$x)-2), 5)
    p |>
      add_trace(
        type = "scatter3d",
        mode = "markers+lines",
        x = ~x,
        y = ~y,
        z = ~z,
        name = col,
        data = d,
        mode = "lines",
        line = list(
          color = d$hex,
          width = 5
        )
        ,marker = list(
          size = sizes,
          symbol =  markers,
          color = d$hex
        )
      ) ->
      p
  }
  p |> 
    layout(title = list(text = mod))->
    p
  if (mode == "dark") {
    p |> 
      layout(
        paper_bgcolor = "#222",
        scene = list(
          xaxis = list(
            domain = c(-1, 1),
            color = "#222"
          ),
          yaxis = list(
            domain = c(-1, 1),
            color = "#222"
          ),
          zaxis = list(
            domain = c(-1, 1),
            color = "#222"
          )      
        ),
        title = list(text = mod, font = list(color = "white"))
      ) ->
      p
  }
  p
}
```

### "Light"

Let's start with "light".
The location of the base color is indicated with a circle, the modified color with a diamond, with a line connecting the two.

::: {.light-content style="width: 100%;"}
```{r}
make_mod_plot(comp_hls, "light")
```
:::

::: {.dark-content style="width: 100%;"}
```{r}
make_mod_plot(comp_hls, "light", mode = "dark")
```
:::

A few cool things here!

-   Even if very dark base colors shot straight upwards to lighten, this would still involve some degree of relative desaturation (they'd be deeper within the sphere).
    As it is, though, it looks like they tend to arc inward.

-   Desaturated colors that started out in the light hemisphere (like "sage" or "rose") look like they make a b-line for the surface of the sphere, which involves *saturating* them.

-   Other highly saturated colors look like they stay close to the saturation surface, rather than cutting through the sphere (and desaturating).

-   A lot of colors look like they've got a bit of a twist towards 90°, greening a little.

### "dark"

::: light-content
```{r}
make_mod_plot(comp_hls, "dark")
```
:::

::: dark-content
```{r}
make_mod_plot(comp_hls, "dark", mode = "dark")
```
:::

A lot of this looks like the reverse pattern of "light."

-   Very light colors really dive downwards and toward the center (desaturating).

-   Relative dark and highly saturated colors stay closer to the surface (keeping their saturation).

### "pale"

::: light-content
```{r}
make_mod_plot(comp_hls, "pale")
```
:::

::: dark-content
```{r}
make_mod_plot(comp_hls, "pale", mode = "dark")
```
:::

This looks really similar to "light".
Maybe the main difference is where the base colors are originating from?
Looks like "pale" doesn't combine with as many colors from the dark hemisphere as "light" did.

### "bright"

::: light-content
```{r}
make_mod_plot(comp_hls, "bright")
```
:::

::: dark-content
```{r}
make_mod_plot(comp_hls, "bright", mode = "dark")
```
:::

This one's really cool: Everything heads towards the equator!

### "deep"

::: light-content
```{r}
make_mod_plot(comp_hls, "deep")
```
:::

::: dark-content
```{r}
make_mod_plot(comp_hls, "deep", mode = "dark")
```
:::

This looks a lot like "dark".
This might be down to there being fewer colors described as "dark \<color\>", but there's no base colors in the light hemisphere between 0° and 270°.

### "dull"

::: light-content
```{r}
make_mod_plot(comp_hls, "dull")
```
:::

::: dark-content
```{r}
make_mod_plot(comp_hls, "dull", mode = "dark")
```
:::

This looks exactly like you'd expect: everything's diving towards the core!

## Wrapping up

I think figuring out all of the math to plot these trajectories through color spheres was totally worth it!
